Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-03-21 paper/finalish/article.tex
move figures to appropriate filenames
use lowercase figures for consistency
adds R script for generating the tables, and update the text accordingly. Just DIC to go then we're done

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrrrr}
      \toprule
      ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm}& \textit{tkt}& \textit{unc}A\\ \midrule
      403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
      2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
      2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The allelic profiles of a selection of genotypes, composed of seven allele numbers at each of the seven loci.}
  \label{tab:mlst}
\end{table}

@@

The source attribution study described here uses human case data ($H=1,400$) collected in the Manawatu region of New Zealand from March 2005 to December 2014. In addition, over the same time period and from the same geographical location, isolates $N=2,080$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others. The total number of unique genotypes from all isolates is $I=355$, with 35\% of genotypes found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases. As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants \cite{Muell, Coll, Cart}. The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.

The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random. Table~\ref{tab:rurality} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}. The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.

\begin{table}
  \begin{center}
    \begin{tabular}{rrrrrr}
      \toprule
      ST & Human & Poultry & Ruminants & Water & Others\\ \midrule
      42 & 54 & 7 & 52 & 10 & 2\\
      45  & 141 & 154 & 10 & 21 & 54\\
      474 & 239 & 60 & 15 & 5 & 7\\
      2026 & 37 & 0 & 46 & 2 & 5\\
      2381 & 0 & 0 & 0 & 21 & 3\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The frequency of five genotypes found from human and four source isolates.}
  \label{tab:counts}
\end{table}

\begin{table}
%  \begin{threeparttable}
  \centering
      \begin{tabular}{rlrrrr}
        \toprule
        Rurality scale & Description & Human cases & 2006 & 2013 \\ \midrule
        -3 & Highly rural/remote area & 19 & 1,572 & 1,527\\
        -2 & Rural area with low urban influence & 92 & 8,373 & 8,304\\
        -1 & Rural area with moderate urban influence & 127 & 10,392 & 10,734\\
        0 & Rural area with high urban influence & 70 & 6,579 & 7,155\\
        1 & Independent urban area & 230 & 28,611 & 28,188\\
        2 & Satellite urban area & 178 & 19,725 & 20,525\\
        3 & Main urban area & 579 & 76,047 & 78,108\\
        \bottomrule
      \end{tabular}
      \caption{The number of human cases in each rurality class from 2005--2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
      \label{tab:rurality}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=.8\linewidth]{Figures/case_rates}
  \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014. An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years, particularly in urban areas.}
  \label{fig:case_rates}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/fig_attribution.pdf}
\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig:attribution}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/genotype_figure.pdf}
\caption{Posterior probability for each source for four sequence types from the asymmetric Island (red) and Dirichlet (blue) models, assuming that each source is \emph{apriori} equally likely.}
\label{fig:genotype_dist}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/fig_intervention.pdf}
\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab:rurality}). The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig:intervention}
\end{figure}

@@

The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model. To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that each source was equally likely. Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources. The asymmetric Island model, however, can exploit the genetic relationship between genotypes. ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables~\ref{tab:mlst} and \ref{tab:counts}). Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources. In contrast, both models provide similar probabilities for ST-2026 and ST-474 which are both observed frequently.

>>>>

\begin{table}
  \begin{center}
	\begin{tabular}{rrrrrrrr}
	\toprule
	ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm} & \textit{tkt} & \textit{unc}A\\
	\midrule
	2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
	474 & 2 & 4 & 1 & 2 & 2 & 1 & 5\\
	2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
	403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
	\bottomrule
	\end{tabular}
  \end{center}
  \caption{The allelic profiles of a selection of genotypes, composed of seven allele numbers at each of the seven loci.}
  \label{tab:mlst}
\end{table}

@@

The source attribution study described here uses human case data ($H=1,460$) collected in the Manawatu region of New Zealand from March 2005 to December 2014. In addition, over the same time period and from the same geographical location, isolates $N=2,128$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others. The total number of unique genotypes from all isolates is $I=348$, with 36\% of genotypes found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases. As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants \cite{Muell, Coll, Cart}. The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.

The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Approximately 8\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random. Table~\ref{tab:rurality} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}. The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.

\begin{table}
  \begin{center}
	\begin{tabular}{rrrrrr}
	\toprule
	ST & Human & Poultry & Ruminants & Water & Others\\
	\midrule
	42 & 59 & 7 & 53 & 10 & 2\\
	45 & 149 & 155 & 10 & 21 & 54\\
	474 & 247 & 60 & 15 & 5 & 9\\
	2026 & 28 & 0 & 40 & 5 & 2\\
	2381 & 0 & 0 & 0 & 60 & 3\\
	\bottomrule
	\end{tabular}
  \end{center}
  \caption{The frequency of five genotypes found from human and four source isolates.}
  \label{tab:counts}
\end{table}

\begin{table}
%  \begin{threeparttable}
  \centering
	\begin{tabular}{rlrrr}
	\toprule
	Rurality scale & Description & Human cases & 2006 & 2013\\
	\midrule
	-3 & Highly rural/remote area & 16 & 1,572 & 1,527\\
	-2 & Rural area with low urban influence & 103 & 8,382 & 8,316\\
	-1 & Rural area with moderate urban influence & 124 & 10,392 & 10,734\\
	0 & Rural area with high urban influence & 78 & 6,579 & 7,155\\
	1 & Independent urban area & 240 & 28,611 & 28,188\\
	2 & Satellite urban area & 187 & 19,725 & 20,526\\
	3 & Main urban area & 596 & 76,047 & 78,108\\
	\bottomrule
	\end{tabular}
      \caption{The number of human cases in each rurality class from 2005--2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
      \label{tab:rurality}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=.8\linewidth]{figures/fig_case_rates.pdf}
  \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014. An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years, particularly in urban areas.}
  \label{fig:case_rates}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{figures/fig_attribution.pdf}
\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig:attribution}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{figures/fig_genotype_dist.pdf}
\caption{Posterior probability for each source for four sequence types from the asymmetric Island (red) and Dirichlet (blue) models, assuming that each source is \emph{apriori} equally likely.}
\label{fig:genotype_dist}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{figures/fig_intervention.pdf}
\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab:rurality}). The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig:intervention}
\end{figure}

@@

The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model. To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that each source was equally likely. Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=6$ cases each) with ST-403 not being observed among the sources, and ST-2343 being observed once in poultry, so that the Dirichlet model has little information available to distinguish between sources. The asymmetric Island model, however, can exploit the genetic relationship between genotypes. ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables~\ref{tab:mlst} and \ref{tab:counts}). Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources. In contrast, both models provide similar probabilities for ST-2026 and ST-474 which are both observed frequently.
